PACKER: An Exemplar Model of Category Generation

### 2023-01-25 Manuscripts/cogsci_2017/paper.tex

PACKER sets $f(x_j)$ depending on exemplar $j$'s category membership: $f(x_j) = \gamma$ if $x_j$ is a member of the target category, and $f(x_j) = 1 - \gamma$ if $x_j$ is a member of a contrast category. $\gamma$ is thus a free parameter ($0 \leq \gamma \leq 1$) controlling the trade-off between within- and between-category similarity. PACKER's core proposal is that new categories should be different from existing categories, and same-category exemplars should be similar to one another. This is realized when $\gamma$ assumes an intermediate value: For example, when $\gamma = 0.5$, $f(x_j) = 0.5$ for members of the target category and $f(x_j) = -0.5$ for members of other categories; thus, the model is likely to generate items that are similar to members of the target category but are not similar to members of other categories. However, more extreme values can be used to produce different behavior, see Figure \ref{fig:example-prob-spaces}.

@@

Previous models of category generation focus on capturing the tendency for people to produce new categories that have similar distributional properties to existing categories. To the best of our knowledge, \citet{jern2013probabilistic} were the first to evaluate computational models of generation. Based on their work, we describe two alternative models: a formalization of the \textit{copy-and-tweak} hypothesis \citep{ward1995s}, and the \textit{hierarchical sampling} hypothesis \citep{jern2013probabilistic}.

\subsection{Copy-and-Tweak}
The \textit{copy-and-tweak} model proposes that generation is a two-part process: First, learners retrieve an observation of the target class from memory, and then they tweak it to make something new. \citet{jern2013probabilistic} interpreted this proposal in terms of an exemplar model using the GCM \citep{nosofsky1984choice}. Formally, their model is equivalent to PACKER with $\gamma = 1$ (see Figure \ref{fig:example-prob-spaces}). In this case,  $f(x_j) = 1$ for members of the target category and $f(x_j) = 0$ for members of other categories; thus, the model considers only target-class similarity, and when no members of the target class are known, the model generates items at random. Thus, in our work we provide simulations from the copy-and-tweak account as a variant of PACKER with a fixed $\gamma$ parameter.

Formalizing a model family where PACKER and copy-and-tweak are different parameterizations within the same framework is useful because comparison between the models thus provides a test of the explanatory value of the contrast mechanism: The account provided by copy-and-tweak will only equal that of PACKER if the contrast mechanism does not offer an advantage (i.e., if $\gamma < 1$ significantly improves model fits).

@@

Each model's best-fitting parameterization is shown in Table \ref{table:model-fits}. Overall, PACKER outperformed copy-and-tweak and the hierarchical sampling model by a considerable margin ($\sim11\%$ improvement in log-likelihood). The parameter settings associated with PACKER's best fit are exactly as expected: a strong preference for items that are similar to members of the target category but are dissimilar to members of the contrast category. A similar pattern of results was obtained when we only considered the second to fourth exemplars generated by each participant.

>>>>

PACKER sets $f(x_j)$ depending on exemplar $j$'s category membership: $f(x_j) = \gamma$ if $x_j$ is a member of the target category, and $f(x_j) = \gamma - 1$ if $x_j$ is a member of a contrast category. $\gamma$ is thus a free parameter ($0 \leq \gamma \leq 1$) controlling the trade-off between within- and between-category similarity. PACKER's core proposal is that new categories should be different from existing categories, and same-category exemplars should be similar to one another. This is realized when $\gamma$ assumes an intermediate value: For example, when $\gamma = 0.5$, $f(x_j) = 0.5$ for members of the target category and $f(x_j) = -0.5$ for members of other categories; thus, the model is likely to generate items that are similar to members of the target category but are not similar to members of other categories. However, more extreme values can be used to produce different behavior, see Figure \ref{fig:example-prob-spaces}.

@@

Previous models of category generation focus on capturing the tendency for people to produce new categories that have similar distributional properties to existing categories. To the best of our knowledge, \citet{jern2013probabilistic} were the first to evaluate computational models of generation. Based on their work, we describe two alternative models: a formalization of the \textit{Path of Least Resistance} hypothesis (later termed {\em copy-and-tweak}, see \citealp{jern2013probabilistic}), and the \textit{hierarchical sampling} hypothesis \citep{jern2013probabilistic}.

\subsection{Copy-and-Tweak}
The \textit{copy-and-tweak} model, based broadly on the earlier Path of Least Resistance view \citep{ward1994structured,ward1995s}, proposes that participants generate categories by retrieving an observation of the target class from memory, and then tweaking it to make something new. \citet{jern2013probabilistic} interpreted this proposal in terms of an exemplar model using the GCM \citep{nosofsky1984choice}. Formally, their model is equivalent to PACKER with $\gamma = 1$ (see Figure \ref{fig:example-prob-spaces}). In this case,  $f(x_j) = 1$ for members of the target category and $f(x_j) = 0$ for members of other categories; thus, the model considers only target-class similarity, and when no members of the target class are known, the model generates items at random.

In our work we provide simulations from the copy-and-tweak account, realized as a variant of PACKER with a fixed $\gamma$ parameter. Formalizing a model family where PACKER and copy-and-tweak are different parameterizations within the same framework is useful because comparison between the models provides a test of the explanatory value of the contrast mechanism: The account provided by copy-and-tweak will only equal that of PACKER if the contrast mechanism does not offer an advantage (i.e., if $\gamma < 1$ significantly improves model fits).

@@

Each model's best-fitting parameterization is shown in Table \ref{table:model-fits}. Overall, PACKER outperformed copy-and-tweak and the hierarchical sampling model by a considerable margin ($\sim13\%$ improvement in log-likelihood). The parameter settings associated with PACKER's best fit are exactly as expected: a strong preference for items that are similar to members of the target category but are dissimilar to members of the contrast category. A similar pattern of results was obtained when we only considered the second to fourth exemplars generated by each participant.
